---
title: "RNAseqEr"
output: rmarkdown::html_vignette
author: "Luise A. Seeker"
date: "25 July 2023"
vignette: >
  %\VignetteIndexEntry{RNAseqEr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,  # Enable caching globally
  fig.width = 8,
  fig.height = 5
)
```

# Fast and reproducible single cell/ nucleus RNAseq analysis

This package was written to allow for a fast and reproducible processing of 
single cell or single nucleus RNA sequencing (sc/snRNAseq) data in R by 
summarizing common workflows into fewer functions and by providing solutions
for bottle-neck steps that often slow down progress such as the cell type annotation
and the decision on a suitable clustering resolution that is biologically meaningful.

The workflow starts with a pre-processed dataset where the following is considered
* genes that are not expressed may be removed from the dataset (gene QC)
* low quality cells are removed from the dataset
* ambient RNA may be removed
* doublets may be removed or flagged
* dataset may be batch corrected and/ or integrated if required.

The workflow consists of
1) Performing standard Seurat processing including
* Normalisation
* finding of variable genes
* linear and non-linear dimensional reductions
* CLustering at different resolutions

2) Finding a stable cluster resolution for automatic celltype annotation using an 
adapted ScType algorithm

3) Subsetting the datasets for cell lineages

4) Finding a suitable clustering resolution for cell lineage datasets (bottleneck)

5) cluster quality control

6) Identification of best cluster marker genes for validation (THIS NEEDS SOME MORE THOUGHT)

7) differential abundance with variables of interest

8) differential gene expression with variables of interest and identification of
  condition markers

9) gene ontology based on cluster and condition markers


10) Building of a shiny app

11) writing of a summary report that describes what has been done, what the 
results are and where to find them. 


The processed Seurat objects are saved which enables the addition of other 
taylored analyses. 

All results should be thoroughly checked and manually corrected if necessary.

# Installation

```{r, eval = FALSE}
# install.packages("devtools")  # Uncomment if devtools is not installed
# devtools::install_github("LASeeker/RNAseqEr")

```


# Load libraries

```{r setup}
library(RNAseqEr)
library(here)
library(Seurat)
library(miloR)
library(SingleCellExperiment)
library(MAST)
library(dplyr)
library(HGNChelper)  # Required for gene symbol checking
library(openxlsx)    # Required for reading Excel files
library(CHOIR)  # Optional package for advanced clustering
```

# Test with provided test dataset or own data

The package comes with a dataset from the human CNS (Seeker et al. 2023) that 
includes different cell types from different CNS regions and different donor 
age and sex groups. This data can be loaded with 

```{r}
cns
```


## Use wrapper function for the entire workflow
```{r, eval = FALSE}

result <- quick_RNAseqEr(cns,
                         cluster_qc_vars = c("process_number", "caseNO", "Tissue"),
                         sample_id = "uniq_id",
                         cols_milo_design = c("uniq_id","Tissue","gender","AgeGroup","caseNO"),
                         milo_test_fact = c("Tissue", "gender", "AgeGroup"),
                         # Volcano plot customization
                         volcano_mode = c("overall", "clusterwise"),
                         volcano_plotheight = 8,
                         volcano_plotwidth = 10,
                         volcano_additional_p_adjust = TRUE,  # Additional p-value adjustment
                         volcano_p_adjust_method = "BH",      # Benjamini-Hochberg method
                         # Shiny app customization
                         shiny_name = "my_analysis",
                         author = "Dr. Smith",
                         title = "My Single Cell Analysis",
                         # Manuscript generation (optional)
                         manuscript_generation = TRUE,
                         manuscript_title = "Single-cell analysis reveals age-related changes in brain tissue",
                         manuscript_author = "Dr. Smith",
                         manuscript_journal = "Nature Communications",
                         manuscript_llm_provider = "openai",
                         manuscript_api_key = "your-api-key-here",  # Set your API key
                         manuscript_model = "gpt-4",
                         manuscript_include_figures = TRUE,
                         manuscript_focus_genes = "condition",
                         manuscript_min_logfc = 0.5,
                         manuscript_max_pvalue = 0.05)
 
```

```{r, eval = FALSE}
#For large datasets, use sketching
result <- quick_RNAseqEr(large_dataset,
                          sketch = TRUE,
                          sketch_ncells = 100000,
                          cluster_qc_vars = c("process_number", "caseNO", "Tissue"),
                          sample_id = "uniq_id",
                          cols_milo_design = c("uniq_id","Tissue","gender","AgeGroup","caseNO"),
                          milo_test_fact = c("Tissue", "gender", "AgeGroup"),
                         # Volcano plot customization
                         volcano_mode = c("overall", "clusterwise"),
                         volcano_plotheight = 8,
                         volcano_plotwidth = 10,
                         volcano_additional_p_adjust = TRUE,  # Additional p-value adjustment
                         volcano_p_adjust_method = "BH",      # Benjamini-Hochberg method
                         # Shiny app customization
                         shiny_name = "my_analysis",
                         author = "Dr. Smith",
                         title = "My Single Cell Analysis")

#Customize Shiny app parameters
result <- quick_RNAseqEr(cns,
                          cluster_qc_vars = c("process_number", "caseNO", "Tissue"),
                          sample_id = "uniq_id",
                          cols_milo_design = c("uniq_id","Tissue","gender","AgeGroup","caseNO"),
                          milo_test_fact = c("Tissue", "gender", "AgeGroup"),
                          # Shiny app customization
                          shiny_name = "my_analysis",
                          default_1 = "AgeGroup",  # Use specific column name
                          default_2 = 5,           # Use 5th metadata column
                          author = "Dr. Smith",
                          title = "My Single Cell Analysis",
                          journal = "Nature",
                          volume = "123",
                          page = "456",
                          year = "2024",
                          doi = "https://doi.org/10.1038/xxx",
                          link = "https://github.com/myself/myproject",
                          shiny_title = "My Amazing Results")

```

## Standard Seurat processing

It can be seen that the cns data above already contains normalized and scaled data as well
as different dimensional reductions. Your data will not have any of those at this
point of the analysis. [Seurat](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) 
is a fantastic and very user-friendly tool for 
adding those information to your data and I highly recommend reading through their 
papers and vignettes. However, if you analyse many datasets, you 
will also see that your work is going to be very repetitive, using the same Seurat 
functions again and again without tweaking them necessarily a lot. So here we 
present a function that automatically runs Seurat's standard normalisation, 
selection of variable genes, dimensional reduction and clustering at different 
resolutions. 

```{r seurat_processing, cache=TRUE, echo=FALSE, results='hide', fig.keep='all', fig.width= 8, fig.height=5}
cns <- seurat_proc(cns)

```

The function plots an elbow plot so that you can check if the standard of 20 principle
components is suitable for your data. If you think you need to use less or more 
(20 is far away from the elbow in the plot), you can re-run the function while
providing an additional argument:

```{r, eval= FALSE}
cns <- seurat_proc(cns, n_pcs = 25)

```


## Plotting of different clustering resolutions

In the previous step you created different clustering resolutions, but which
is the most appropriate one for your data? This is often a tricky question to 
answer and the literature on this is vast. 

A first good step is to visualise different clustering resolutions. 
You can easily do this with the function below which will also save dimensional 
reduction plots (default is umap, tsne can be used with use_reduction = "umap") 
to a folder. Remember to set your save_directory that you'd like to use, the 
default is the current working directory. 


```{r,  echo=FALSE,results='hide', fig.keep='all', fig.width= 8, fig.height=5}

plot_list(cns, save_dir = "../")

```


## Calculate cluster purity at different resolutions

### RNAseqEr approach based on purity measures

The intention above was to create clustering from a too low to a too high resolution. 
But which one is appropriate? This depends in this work for on if you have a 
very homogeneous cell population or a mix of different cell types (which is the
case here), where we would like to do first a rough annotation.

If the second is true, we will use cluster purity measures to first separate the
dataset best for clusters that are physically distinct on in the umap/tsne space 
and thus most likely represent different cell types.

```{r,  echo = FALSE, results = 'hide', fig.keep = 'all', fig.width= 8, fig.height=5}
clu_pure(cns, save_dir = "../")

```

```{r, RNAseqErh = 8, fig.height = 5}
keep_res <- max_pure(save_dir = "../", dir_lab = "all_celltypes")

DimPlot(cns, group.by = keep_res, cols = colour_palette())

```

### Using CHOIR
A recent publication by Sant et al. proposes 

Sant, C., Mucke, L. & Corces, M.R. CHOIR improves significance-based detection of cell types and states from single-cell data. Nat Genet (2025). https://doi.org/10.1038/s41588-025-02148-8

```{r, eval = FALSE}
cns_choir <- CHOIR(cns)
cns_choir <- runCHOIRumap(cns_choir,
                       reduction = "P0_reduction")

plotCHOIR(cns_choir)    
plotCHOIR(cns_choir,
          accuracy_scores = TRUE,
          plot_nearest = FALSE)
```

## Cell type annotation

Above we found a cluster resolution that separates what could be celltypes, however, 
we don't know which cell types are present. We are therefore using an annotation
based on [ScType](https://www.nature.com/articles/s41467-022-28803-w). We are looking 
for a rough annotation, that can be used for subsetting the dataset into lineages 
for subsequent further clustering. 
SCType sometimes returns "Unknown" and we therefore adapted it for those cases 
to not only consider gene expression but also differential gene expression. 
Running a differential gene expression analysis at this step is time consuming 
but also valuable as it will help to double check the automatic annotation. 
The differential gene expression results are saved to file and can be used
for plotting heatmaps for example that show cell type segregation. 

N.B. The SCType and RNAseqER annotation will look identical as long as there are
no "Unknown" in the former. 

TO DO ADD OVERALL DOTPLOT!



```{r, fig.width = 8, fig.height = 5, echo = FALSE, results = 'hide', fig.keep = 'all',}
here("..")
cns <- annotate_seqEr(cns, ident_use = keep_res, tissue_ref = "Brain",
                      save_dir = "../")


cns <- annotate_seqEr(cns, ident_use = "RNA_snn_res.0.8", tissue_ref = "Brain",
                      save_dir = "../")

```

The annotation above looks very close to the manual annotation we used in 
Seeker et al. 2023:

```{r, fig.width = 8, fig.height = 5}
DimPlot(cns, group.by = "rough_annot", cols = colour_palette(), label = TRUE)

```


## Finer annotation

We may or may not be interested in all celltypes that a dataset contains. In many
cases is is beneficial to subset the dataset into cell lineage datasets that then
can be used for finer clustering and downstream analyses. 
The following function subsets the dataset for all clusters that
have been identifies and annotated above using the RNAseqEr annotation. Any 
metadata column can be used for subsetting. Each cell lineage will be saved in 
a separate .RDS file.


```{r}
here("..")
subset_RNAseqEr(cns, save_dir = here())

list.files(here("outs", "data"))


```

## Finer clustering

Finding an appropriate clustering resulution within a cell lineage is often a 
step that takes a long time and is associated with insecurity, particularly when 
the data scientist is yet inexperienced. From a biological point of view 
we seek to find a resolution with the maximum number of clusters that still allows
the identification of cluster markers that can be used for validation using other 
lab techniques such as immuno-fluorescence or RNAscope. 
We show here our approach using the example of ologodendrocytes. 
We also provide a function using the same approach for all the subsetted datasets
automatically. 

```{r, fig.width = 10, fig.height = 5, echo = FALSE, results = 'hide', fig.keep = 'all'}

seur <- readRDS(here("outs", "data", "Oligodendrocytes.RDS"))
seur <- seurat_proc(seur, tsne = FALSE)

```


```{r, fig.show ='hold', out.width ='50%', echo = FALSE,results ='hide', fig.keep ='all'}
plot_list(seur, save_dir = "../", dir_lab = "Oligodendrocytes")

```


In the above plots it can be seen that some resolutions do not seem to make sense
as they either don't cluster at all (< 0.1), have the same clusters as another
resolution (for example 0.6 and 0.7) or simply seem to overcluster, where cluster
separations are visually not detectable and cluster labels overlap completely (>1).

All other cluster resolutions may make sense, so how to pick a suitable one? Our
approach is to decide in by looking at the results of differential gene expression 
analyses to see whether or not we can identify markers for most clusters. Why
most and not all? We noticed that in many datasets, one cluster seems to be the 
"average" that does express all the cell lineage markers but nothing special over 
and above. It may be reasonable to keep a clustering resolution where a cluster
is negative for markers that characterise other clusters.

So all potentially interesting resolutions can be used below:

```{r}

# Pick interesting resolutions
int_resol <- c(0.1, 0.2, 0.3, 1, 1.3)
int_res <- paste0("RNA_snn_res.", int_resol)

```


```{r, eval = TRUE, echo = FALSE, results = 'hide'}
            
all_res_mark <- int_res_all_mark(seur_obj = seur,
                                 int_cols = int_res,
                                 save_dir = "../",
                                 dir_lab = "Oligodendrocytes")

```


```{r}
# see the output 
list.files(here("outs", "Oligodendrocytes", "tables", "cluster_marker", "overall"))
```

The function above generates two outputs for each resolution, one set for unfiltered ("all")
and one for further filtered ("fil") values that can be tweaked in the function arguments.

We also add a pairwise comparison of clusters:



```{r, eval = TRUE, echo = FALSE, results = 'hide'}

pairwise_dge(seur_obj = seur,
             int_cols = int_res,
             save_dir = "../",
             dir_lab = "Oligodendrocytes")
list.files(here("outs", "Oligodendrocytes", "tables", "cluster_marker", "pairwise"))
```


Above we generates a whole lot of output that nobody wants to comb through 
manually. So below we are going to read in all the results and summarise them
in a meaningful way. 

The two functions below look into the results of the overall (first) and pairwise
(second) differential gene expression analysis results and keeps the top n 
markers (default n = 10) per cluster. 

Those genes are then concatenated and duplicates are filtered out, leaving the
user with a condensed list of the most interesting potential cluster markers which
are used to determine the similarity/ difference of clusters to find an appropriate
clustering resolution.

```{r}
# Check if files exist before trying to read them
overall_dir <- here("outs", "Oligodendrocytes", "tables", "cluster_marker", "overall")
pairwise_dir <- here("outs", "Oligodendrocytes", "tables", "cluster_marker", "pairwise")

if(dir.exists(overall_dir) && length(list.files(overall_dir)) > 0) {
  clu_mark <- gen_mark_list(file_dir = overall_dir)
} else {
  clu_mark <- data.frame(gene = character(), stringsAsFactors = FALSE)
  cat("No overall marker files found. Using empty data frame.\n")
}

if(dir.exists(pairwise_dir) && length(list.files(pairwise_dir)) > 0) {
  clu_mark_pw <- gen_mark_list(file_dir = pairwise_dir, pairwise = TRUE)
} else {
  clu_mark_pw <- data.frame(gene = character(), stringsAsFactors = FALSE)
  cat("No pairwise marker files found. Using empty data frame.\n")
}

int_genes <- c(clu_mark$gene, clu_mark_pw$gene)
int_genes <- unique(int_genes)
```


## Identifying clusters that are too similar to find most appropriate resolution

Below we use heatmaps to visually screen differences and similarities between
clusters. The function also uses average expression values to compare each 
cluster pair at different reolutions paorwise. If they are too similar based on
their maximum difference, mean difference or Eucledian distance, it's recommended
to merge clusters. Each of those incidences adds 1 to a cluster similarity score. 
If a resolution has a high overall cluster similarity score it means that resolution
includes a lot of clusters that are very similar. The cluster resolution with the 
lowest cluster similarity score is the one the user should consider. 

```{r,  fig.height = 8, fig.width = 8, echo = FALSE, results = 'hide', fig.keep = 'all'}

hm_dir <- here("outs",
               "Oligodendrocytes",
               "plots",
               "heatmaps")

dir.create(hm_dir, recursive = TRUE)

# Only run heatmap if we have genes to plot
if(length(int_genes) > 0) {
  df_seqEr <- heatmap_seqEr(seur,
                use_resol = FALSE,
                col_names = int_res,
                int_genes = int_genes,
                save_dir = hm_dir,
                max_diff_threshold = 10,
                mean_diff_thres = 0.1)
} else {
  df_seqEr <- data.frame(resolution = "No genes available", cluster_sim_score = 0)
  cat("No genes available for heatmap analysis.\n")
}
```

To look at the cluster similarity scores, inspect the dataframe below. It shows that
in this case the resolution 1 may be the most appropriate. More generally we use the 
highest resolution with the lowest cluster similarity score (tail(x,1)).

```{r}
df_seqEr
```

The below only works if there was some over-clustering and there is indeed one cluster
with a minimum cluster similarity score. Alternatively this can be determined manually.

```{r, eval = FALSE}
res_df <- subset(df_seqEr, df_seqEr$cluster_sim_score == min(df_seqEr$cluster_sim_score))
keep_res <- tail(res_df$resolution,1)
```

```{r}
keep_res <- "RNA_snn_res.1"

```


```{r, fig.width = 8, fig.height = 5}
Idents(seur) <- keep_res

seur$cluster_id <- paste0("Oligodendrocytes_", seur@meta.data[[keep_res]])

DimPlot(seur, group.by = "cluster_id", label = TRUE, cols = colour_palette())
```

## Cluster Quality control

Sometimes individual samples can cluster separately from the rest which is 
more likely to be due to technical reasons than showing a biologically interesting
signal. This may particularly happen when working with human or other non-model
organism data where the post-mortem conditions are more variable and there is more
genetic variation and variation in lifestyle, disease burden and medication. 

We always perform some form of cluster quality control to capture clusters that
are made up only by a few samples or donors. If tiny cluster are as a consequence 
removed from the dataset, we keep the rest of the dataset as it is. If a significant 
proportion is removed, it is a safer option to then repeat the selection of variable 
genes up to the re-clustering. 

The function below can be used on any meta data column to test for cluster
biased clustering. 

```{r, fig.width = 8, fig.height = 5}
seur <- cluster_qc(seur_obj = seur,
                   save_dir = save_dir,
                   cluster_col = "cluster_id",
                   vars = c("process_number", "caseNO", "Tissue"),
                   dir_lab = "Oligodendrocytes")


```


# Differential abundance testing

Often it is of interest to see if the number of cell/ nuclei per cluster vary 
depending on the condition for example. 
Different methods have been implemented to test this question in a way that is 
statistically sound which is less trivial then it may appear at first. 
In our hands Milo (Ref) works well and we implemnted it here. We encourage the
user to try and test other methods as well. 

```{r, fig.width= 8, fig.height=5}
milo_obj <- abundance_test(seur_obj = seur,
                           cluster_col = "cluster_id",
                           sample_id = "uniq_id",
                           cols_interest = c("uniq_id",
                                             #"Tissue",
                                             #"gender",
                                             "AgeGroup" #,
                                             #"caseNO"
                                             ),
                           test_factors = c(#"Tissue", 
                                            #"gender", 
                                            "AgeGroup"),
                           dir_lab = "Oligodendrocytes",
                           k = 10,           # Reduced from 30
                           d = 10,           # Reduced from 30
                           prop = 0.1,       # Reduced from 0.2
                           refined = FALSE  #
                           )

```
Differential abundance testing makes more sense for larger datasets. Therefore,
I am showing it below for the cns dataset containing all  cell types.

```{r}
milo_obj <- abundance_test(seur_obj = cns,
                       cluster_col = "rough_annot",
                       sample_id = "uniq_id",
                       cols_interest = c("uniq_id",
                                          #"Tissue",
                                          #"gender",
                                          "AgeGroup"#,
                                          #"caseNO"
                                          ),
                      test_factors = c(#"Tissue", 
                                       #"gender", 
                                       "AgeGroup"
                                       ),
                      save_dir = "../",
                           k = 10,           # Reduced from 30
                           d = 10,           # Reduced from 30
                           prop = 0.1,       # Reduced from 0.2
                           refined = FALSE  #
                      )


```

# Differential gene expression with condition

Next it is interesting to find out which genes are differentially expressed with
variables of interest such as disease condition, treatment group, age group or
tissue region for example. One approach is to see which genes are expressed in 
the clusters more abundant in those groups (see above). The other is to look
within cell lineage or clusterwise for genes that are differentially expressed.



```{r}

RNAseqEr::find_cond_markers(seur,
                  int_cols = c("AgeGroup","Tissue", "gender"),
                  cluster_id = "cluster_id",
                  dir_lab = "Oligodendrocytes",
                  save_dir = "../")
            

```
The subsetted oligodendrocyte dataset is very small to test for differentially 
expressed genes. This is why we show the workflow downstream with the cns dataset.


```{r}

RNAseqEr::find_cond_markers(cns,
                  int_cols = c("AgeGroup","Tissue", "gender"),
                  cluster_id = "rough_annot",
                  save_dir = "../")
            

```

The example below shows how to collect data from the cluster-wise differential gene
expression analysis that we performed above and plot them as violin plots. 
```{r, fig.height= 50, fig.width=8}
library(here)
cond_genes <- gen_mark_list(file_dir = here("outs", 
                                            "all_celltypes", 
                                            "tables", 
                                            "condition_mark" , 
                                            "rough_annot" , 
                                            "clusterwise"),
                            condition = TRUE,
                            test_cond = c("AgeGroup", "Tissue", "gender"))

# Subset for the condition we would like to plot starting with age
age_mark <- subset(cond_genes, 
                              cond_genes$condition == "AgeGroup")

#There is no point in plotting the same genes several times. Therefore, we are
# removing suplicated gene names.

age_mark <- age_mark[!duplicated(age_mark$gene),]
head(age_mark)

# Save vln plots to file
save_vln(cns, 
         marker_list = age_mark$gene,
         save_label = "Age_mark",
         split_by = "AgeGroup",
         group_by = "rough_annot",
         plotheight = 50,
         save_dir = "../")



```

The same is also possible to perform on the overall differential gene expression
analysis from above by pointing at the corresponding folder.

```{r, plot.width = 8, plot.height = 30}
cond_genes_ov <- gen_mark_list(file_dir = here("outs", 
                                            "all_celltypes", 
                                            "tables", 
                                            "condition_mark" , 
                                            "rough_annot" , 
                                            "overall"),
                            condition = TRUE,
                            test_cond = c("AgeGroup", "Tissue"))

# Subset for the condition we would like to plot, this time choosing "Tissue"
tissue_mark <- subset(cond_genes_ov, 
                              cond_genes_ov$condition == "Tissue")

#There is no point in plotting the same genes several times. Therefore, we are
# removing suplicated gene names.

tissue_mark <- tissue_mark[!duplicated(tissue_mark$gene),]
head(tissue_mark)

# Save vln plots to file
save_vln(cns, 
         marker_list = tissue_mark$gene,
         save_label = "Tissue_mark",
         split_by = "Tissue",
         group_by = "rough_annot",
         plotheight = 30,
         save_dir = "../")




```
Similarly, Feature plots can be useful to look visualize differences in gene expression. 
We can gene list above to visualize differences with a condition group such as tissue.


```{r, plot.width = 8, plot.height = 30}
save_feat_plots(cns, 
                marker_list = tissue_mark$gene,
                save_label = "Tissue_mark",
                split_by = "Tissue",
                plotheight = 30,
                save_dir = "../")


```

# Volcano plots for condition markers

Volcano plots are excellent for visualizing differential gene expression results, 
showing the relationship between statistical significance (p-value) and biological 
relevance (fold change). The `volcano_seqEr` function can create volcano plots 
that show multiple comparisons in a single plot, making it easy to identify 
genes that are differentially expressed across different conditions.

The function can work in two modes:
- **"overall" mode**: Shows all cell types together in one plot
- **"clusterwise" mode**: Creates separate plots for each cell type

```{r, eval = FALSE}
# Generate volcano plots for condition markers
volcano_seqEr(dge_res = cond_genes,
              clu_col = "cluster",
              condition_label = "condition",
              mode = "overall",  # or "clusterwise"
              save_dir = "../",
              conditions = c("AgeGroup", "Tissue"),
              plotheight = 6,
              plotwidth = 8,
              additional_p_adjust = TRUE,  # Additional p-value adjustment across all comparisons
              p_adjust_method = "BH")     # Benjamini-Hochberg method
```

The volcano plots will be saved in the `outs/summary_figures/conditions/volcano/` 
directory and will show:
- Genes with high fold change and significance highlighted
- Color coding by cell type
- Top 5 most significant genes labeled
- Clear separation between upregulated and downregulated genes

**Note on Multiple Testing Correction:**
The function uses Seurat's adjusted p-values (`p_val_adj`) by default, which applies 
Bonferroni correction within each comparison. For additional rigor when testing 
multiple conditions or cell types, you can enable `additional_p_adjust = TRUE` 
to perform an additional p-value adjustment across all comparisons using methods 
like Benjamini-Hochberg (BH) or others.

# Gene ontology analysis

It may also be useful to check whether there are certain pathways differentially 
regulated in certain clusters or with a certain condition. There are many 
tools available for gene ontology or gene set enrichment analyses and even more
ways to visualise the results. We use ClusterProfiler below and generate an output 
that can be easily plotted and saved as wished. 

Please note that the usage below is just for the purpose of demonstrating the 
function. When you use the function for your purposes, make sure to filter 
your list of genes so that you retain only those that are significantly upregulated
in your group of interest. Run the function for each condition/
cluster. 

```{r}
go_results <- perform_go(cns,
                         gene_list = tissue_mark,
                         min_log2FC = 0.25,
                         reverse = FALSE,
                         translate_gene_id_from = "SYMBOL",
                         translate_gene_id_to = "ENTREZID",
                         org_use = "org.Hs.eg.db",
                         ontology = "BP",
                         pvalue_cutoff = 0.05,
                         qvalue_cutoff = 0.05,
                         read_able = TRUE)

dotplot(go_results)

```
#Generation of a Shiny app

```{r}
create_shiny(cns,
             shiny_name = "all_celltypes",
             save_dir = "../",
             read_file = FALSE,
             default_1 = "AgeGroup",
             default_2 = "Tissue",
             assay_use = "RNA",
             gex_slot = c("data", "scale.data", "counts"),
             gene_mapping = FALSE,
             default_gene1 = "MALAT1",
             default_gene2 = "GAPDH",
             default_multigene = NA,
             default_dimred = c("umap1", "umap2"),
             author = "My name",
             title = "My data shows amazing results",
             journal = "Best journal I could imaging",
             volume  = "101",
             page    = "101",
             year    = "curren year",
             doi     = "TBC",
             link    = "TBC",
             shiny_title = "My Single Cell Experiment")

```

This function will generate a directory called "shiny_app" in the provided save_dir. 
In that directory is a script called ui.R. When this script is opened in R Studio, 
an option to "Run App" will appear on the top right. When this is clicked, a website
will be opened that can published to share data with collaborators (secret link) and/
or the world. Several datasets can be visualized in the same shiny app as shown in 
https://seeker-science.shinyapps.io/shiny_app_multi/ for example.
More data visualization modes are available in the drop down menu of each dataset/
cell type tab.

## Manuscript Generation with LLM Integration

RNAseqEr can optionally generate manuscript drafts using LLM integration. This feature 
analyzes your results and creates a first draft of a scientific manuscript, focusing 
on the most biologically relevant findings.

### Key Features:

**Automatic Analysis:**
- Identifies top differentially expressed genes
- Focuses on condition-specific changes
- Selects most informative plots for figures
- Generates summary statistics

**LLM Integration:**
- Supports OpenAI GPT-4, Anthropic Claude, and local LLMs
- Customizable journal style (Nature, Cell, bioRxiv, etc.)
- Configurable focus on specific gene categories
- Adjustable statistical thresholds

**Output Files:**
- `manuscript_draft.md`: Complete manuscript draft
- `figure_selection.md`: Recommended plots for figures
- `summary_statistics.md`: Key statistics and results

### Usage Example:

```{r, eval = FALSE}
# Enable manuscript generation
result <- quick_RNAseqEr(cns,
                         # ... other parameters ...
                         manuscript_generation = TRUE,
                         manuscript_title = "Single-cell analysis reveals age-related changes in brain tissue",
                         manuscript_author = "Dr. Smith",
                         manuscript_journal = "Nature Communications",
                         manuscript_llm_provider = "openai",
                         manuscript_api_key = "your-api-key-here",
                         manuscript_model = "gpt-4",
                         manuscript_include_figures = TRUE,
                         manuscript_focus_genes = "condition",
                         manuscript_min_logfc = 0.5,
                         manuscript_max_pvalue = 0.05)
```

### Important Notes:

**Purpose:** The generated manuscript serves as a starting point for:
- Validating results through manual analysis
- Planning laboratory experiments for validation
- Identifying key genes for follow-up studies
- Understanding the biological significance of findings

**Validation Required:** The LLM-generated manuscript should be:
- Reviewed by domain experts
- Validated through independent analysis
- Supplemented with experimental validation
- Refined based on peer review feedback

**API Requirements:** To use LLM integration, you'll need:
- OpenAI API key (for GPT-4)
- Anthropic API key (for Claude)
- Or access to a local LLM service

**Environment Setup:** Create a `.env` file in your project root with your API keys:
```
API_KEY_GPT4=your-openai-api-key-here
ANTHROPIC_API_KEY=your-anthropic-api-key-here
```

## Multi-Species Support

RNAseqEr supports analysis of single-cell RNA-seq data from multiple species. 
The package automatically adapts to different species by:

### **Supported Species:**
- **Mammals:** Human, Mouse, Rat, Cow, Pig, Dog, Cat, Horse, Sheep, Goat, Rabbit, Guinea Pig, Hamster
- **Primates:** Monkey, Chimpanzee, Gorilla, Orangutan, Marmoset, Squirrel Monkey, Tarsier, Bushbaby
- **Other:** Zebrafish, Drosophila, Chicken

### **Automatic Adaptations:**
- **Gene Ontology Databases:** Automatically selects appropriate organism database
- **Tissue References:** Adapts ScType annotation for different tissues
- **Gene ID Mapping:** Handles species-specific gene identifiers
- **Statistical Methods:** All statistical methods work across species

### **Usage Examples:**

```{r, eval = FALSE}
# Human brain analysis (default)
result <- quick_RNAseqEr(human_data,
                         species = "Human",
                         tissue_ref_annotation = "Brain")

# Mouse liver analysis
result <- quick_RNAseqEr(mouse_data,
                         species = "Mouse",
                         tissue_ref_annotation = "Liver")

# Zebrafish brain analysis
result <- quick_RNAseqEr(zebrafish_data,
                         species = "Zebrafish",
                         tissue_ref_annotation = "Brain")

# Drosophila analysis
result <- quick_RNAseqEr(drosophila_data,
                         species = "Drosophila",
                         tissue_ref_annotation = "Brain")
```

### **Species-Specific Functions:**

```{r, eval = FALSE}
# Check if species is supported
is_species_supported("Mouse")  # Returns TRUE
is_species_supported("Alien")  # Returns FALSE

# Get organism database for gene ontology
get_organism_db("Mouse")       # Returns "org.Mm.eg.db"
get_organism_db("Zebrafish")   # Returns "org.Dr.eg.db"

# Get species-specific parameters
params <- get_species_params("Mouse", "Liver")
```

### **Important Notes:**
- **Gene Symbols:** Ensure your data uses appropriate gene symbols for the species
- **Tissue Compatibility:** Some tissues may have limited annotation support for certain species
- **Database Installation:** Install appropriate organism databases (e.g., `org.Mm.eg.db` for mouse)
- **Validation:** Always validate results with species-specific literature

## Environment Setup for LLM Integration

RNAseqEr can automatically load API keys from environment variables, making it easy to use LLM integration without hardcoding sensitive information.

### **Setting Up Your Environment:**

#### **Option 1: Using .env file with dotenv package (recommended for production)**

1. **Install the dotenv package:**
   ```{r, eval = FALSE}
   install.packages("dotenv")
   ```

2. **Create a `.env` file** in your project root directory:
   ```
   API_KEY_GPT4=your-openai-api-key-here
   ANTHROPIC_API_KEY=your-anthropic-api-key-here
   ```

3. **Test your setup:**
   ```{r, eval = FALSE}
   # Test environment setup (automatically installs dotenv if needed)
   setup_rnaseqer_env()
   
   # Check available providers
   check_api_keys()
   
   # Test API key retrieval
   get_api_key(provider = "openai")
   ```

#### **Option 2: Manual setup (for testing)**

If you have issues with the .env file, you can set the API key manually:

```{r, eval = FALSE}
# Set API key manually
set_api_key_manual("your-api-key-here", "openai")

# Test the setup
test_api_key_setup()
```

3. **Use in analysis:**
   ```{r, eval = FALSE}
   # The workflow will automatically use your API key
   result <- quick_RNAseqEr(data,
                            manuscript_generation = TRUE,
                            manuscript_llm_provider = "openai")
   ```

### **Supported Environment Variables:**
- `API_KEY_GPT4` - OpenAI API key
- `OPENAI_API_KEY` - Alternative OpenAI key name
- `ANTHROPIC_API_KEY` - Anthropic API key
- `CLAUDE_API_KEY` - Alternative Anthropic key name


# Output Structure

RNAseqEr creates a comprehensive output structure organized in the `outs/` directory. Here's the complete folder structure:

```
outs/
├── all_celltypes/                    # Main analysis results
│   ├── plots/
│   │   ├── resolution_plots/         # UMAP plots at different resolutions
│   │   ├── final_clustering/         # Final clustering plots
│   │   ├── Condition/                # Condition-specific plots
│   │   │   ├── AgeGroup_mark/
│   │   │   │   └── violin_plots/
│   │   │   └── Tissue_mark/
│   │   │       └── violin_plots/
│   │   └── heatmaps/                 # Cluster similarity heatmaps
│   ├── tables/
│   │   ├── DGE/
│   │   │   └── broad_celltype_markers/  # Differential gene expression results
│   │   ├── condition_mark/
│   │   │   └── RNAseqEr_annotation/
│   │   │       ├── clusterwise/      # Cluster-wise condition markers
│   │   │       └── overall/          # Overall condition markers
│   │   └── cluster_marker/
│   │       ├── overall/              # Overall cluster markers
│   │       └── pairwise/             # Pairwise cluster comparisons
│   └── summary_figures/
│       └── conditions/
│           └── volcano/               # Volcano plots for condition markers
├── data/                             # Processed Seurat objects
│   ├── processed/
│   │   └── all_data/                 # Main processed dataset
│   ├── Astrocytes.RDS                # Subsetted cell lineage data
│   ├── Oligodendrocytes.RDS
│   ├── Neurons.RDS
│   └── ...
├── supplementary_tables/              # Summary tables
│   ├── overall_condition_markers.csv
│   └── clusterwise_condition_markers.csv
├── shiny_app/                        # Interactive Shiny application
│   ├── ui.R
│   ├── server.R
│   └── www/
└── reports/                          # Analysis reports
    ├── methods_report.md
    ├── decision_justifications.md
    └── complete_analysis_report.md
```

### Key Output Files:

**Plots:**
- `resolution_plots/`: UMAP plots at different clustering resolutions
- `final_clustering/`: Final clustering plots for each cell lineage
- `Condition/`: Violin plots and feature plots for condition markers
- `heatmaps/`: Cluster similarity heatmaps
- `volcano/`: Volcano plots showing differential expression

**Tables:**
- `DGE/`: Differential gene expression results
- `condition_mark/`: Condition-specific marker genes
- `cluster_marker/`: Cluster-specific marker genes
- `supplementary_tables/`: Summary CSV files

**Data:**
- `.RDS` files: Processed Seurat objects for each cell lineage
- `processed/`: Main processed dataset

**Applications:**
- `shiny_app/`: Interactive web application for data exploration
- `reports/`: Comprehensive analysis reports

# Session Info

```{r}
sessionInfo()

```
